---
title: "Assignment 2: Coding Basics"
author: "Andrea Gonzalez Natera"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.
```{r}
Seq1 <- seq(1,100,4)
Seq1
```


2. Compute the mean and median of this sequence. 
```{r}
mean(Seq1)
median(Seq1)
```

3. Ask R to determine whether the mean is greater than the median. 
``` {R}
mean(Seq1)>median(Seq1)
```

4. Insert comments in your code to describe what you are doing.

```{r}
#1. I created a sequence from 1 to 100 in intervals of 4  using the seq() function and called this Seq1

#2. Used the mean() and median() functions on Seq1 

#3. Determined if mean of Seq1 was bigger than median by using the > sign

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.
``` {R}
sName <- c('Ale','Sabine','Mariana','Andrea')
#sName is a character vector
sScore <- c(90,85,50,95)
#sScore is a numeric vector
sPassed <- c(TRUE,TRUE,FALSE,TRUE)
#sPassed is a logical vector

```

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{R}
TestResults <- data.frame(sName,sScore,sPassed)
TestResults
```
8. Label the columns of your data frame with informative titles.

```{r}
TestResults <- data.frame(Name = sName, Score = sScore, Result = sPassed)
TestResults
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: This dataframe has three different types of data (numeric,character,logistic) a matrix can only have one type of data

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.
```{R}
PassFail <- function(x){
  ifelse(x>50,TRUE,FALSE)
  }

```

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
PassFail(sScore)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: I used ifelse and it worked because it evaluates every vector item, if I try to use if and else I get an error saying that only the first element in the condition will be used. 
